---
title: "Choosing the Number of Topics"
subtitle: "Model Selection, Evaluation, and Stability"
author: "by Francesco Grossetti"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing the Number of Topics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
has_topicmodels <- requireNamespace("topicmodels", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_topicmodels
)
```

Choosing the number of topics, `k`, is the hardest decision in applied topic
modeling. It is a modeling choice, not a software argument: too few topics blur
distinct themes together, too many split a single theme into near-duplicates, and
there is no single oracle metric that settles the question on its own. The honest
workflow is to fit a grid of candidate models, look at several complementary
pieces of evidence, and check that the chosen model is stable enough to support
the interpretation you intend to publish.

This vignette shows how NLPstudio supports that workflow end to end. It assumes
you have already read the [Topic Model API and Usage](topic-model-api.html)
article, which introduces `fit_topic_model()`, the standardized DTW/TWW outputs,
and the evaluation verbs. Here we focus specifically on selecting `k`: fitting a
candidate grid, reading the selection evidence, evaluating a chosen model in
depth, assessing seed stability, and running a formal statistical test for the
optimal topic count.

For a principled statistical answer to the question, NLPstudio is built to
interoperate with the **OpTop** package and the chi-square optimal-topic test of
Lewis and Grossetti (2022), published in the *Journal of Machine Learning
Research*. That test is treated as a first-class part of the selection toolkit
and is covered in its own section below.

To keep package checks fast and reliable, only the **topicmodels** examples are
evaluated. The OpTop section is shown as a non-evaluated template because OpTop is
an external package.

## A Real Corpus: Presidential Inaugural Addresses

The examples use a real, public corpus that ships with **quanteda**:
US presidential inaugural addresses. We restrict attention to the modern
presidency (1933 onward) so the grid fits quickly while still giving the model
enough material to find recognizable themes such as economic policy, national
security, and civic values.

Corpus preparation is covered in detail in the
[Corpus Preparation and Text Analysis](corpus-workflow.html) article, so here we
keep it compact: tokenize, lowercase, drop stopwords and very short tokens, and
trim rare features so the vocabulary is stable across the candidate models.

```{r corpus}
library(NLPstudio)
library(quanteda)
library(data.table)

inaugural <- corpus_subset(data_corpus_inaugural, Year >= 1933)

toks <- tokens(
  inaugural,
  remove_punct = TRUE,
  remove_numbers = TRUE,
  remove_symbols = TRUE
)
toks <- tokens_tolower(toks)
toks <- tokens_remove(toks, pattern = stopwords("en"))
toks <- tokens_select(toks, min_nchar = 3)

dfmat <- dfm(toks)
dfmat <- dfm_trim(dfmat, min_termfreq = 5, min_docfreq = 2)

dfmat
```

## What "Choosing K" Means

Model-selection evidence for topic models comes from a few families of metrics,
and each one captures a different facet of quality:

- **Coherence** (`coherence_umass`, `coherence_npmi`) rewards models whose top
  terms co-occur in the corpus. Higher coherence usually means more readable
  topics.
- **Exclusivity** rewards top terms that are distinctive to one topic rather than
  shared across many. It guards against the failure mode where every topic is
  dominated by the same generic words.
- **Diversity** measures how much the top-term sets overlap across topics. Low
  diversity is a warning sign of redundant topics, which often appears when `k`
  is too large.
- **Held-out perplexity** (`held_out_perplexity`) measures predictive fit on
  documents the model did not see during training. It is a likelihood-based view
  and frequently disagrees with the interpretability metrics, which is exactly
  why it is worth looking at several at once.

No single number is decisive. Coherence and predictive fit can point in opposite
directions, and the right `k` is the one that best serves the research question.
For a formal decision rule rather than a visual judgment call, the chi-square
test of Lewis and Grossetti (2022) provides a statistical criterion for the
optimal topic count; see the OpTop section below.

## Fit a Grid of Candidate Models

`select_k_topics()` fits one model per value of `k_grid` and returns the
requested metrics in a single long-format selection table. We hold out a portion
of the corpus so that `held_out_perplexity` is available alongside the
interpretability metrics. This teaching corpus is deliberately small, so the
held-out shard is small too; `select_k_topics()` warns about this, and the
predictive numbers below should be read as indicative rather than precise. On a
research-scale corpus the held-out metrics become far more stable.

```{r select-k, warning=FALSE}
k_selection <- select_k_topics(
  dfmat,
  engine  = "topicmodels",
  model   = "lda",
  method  = "Gibbs",
  k_grid  = 4:8,
  metrics = c("coherence_umass", "diversity", "exclusivity",
              "held_out_perplexity"),
  top_n   = 10L,
  holdout = 0.2,
  seed    = 42L,
  control = list(fit = list(iter = 400L, burnin = 100L, thin = 1L))
)

k_selection
```

## Read the Selection Evidence

The selection table follows the package-standard long schema. Every row carries
the same columns regardless of which metric or engine produced it:

- `metric` -- the metric name.
- `level` -- `aggregate` for whole-model metrics, `topic` for per-topic metrics.
- `topic_id` -- the standardized topic identifier, or `NA` for aggregate rows.
- `value` -- the metric value.
- `supported` -- whether the metric is available for this engine and model.

Aggregate rows use `topic_id = NA` rather than dropping the column, which keeps
joins, plots, and exports predictable. The `k` column identifies the candidate
model.

For a paper or appendix you usually want one row per candidate `k` with the
aggregate evidence side by side. `summarize_k_selection()` builds that wide
report and preserves the topic-level rows as an attribute when they are present.

```{r selection-summary}
k_report <- summarize_k_selection(k_selection)
k_report
```

## Visualize the Trade-offs

A plot often makes the trade-offs easier to read than a table. The `plot()`
method for the selection object draws each aggregate metric against `k`, faceted
by metric, so you can see where coherence peaks, where diversity starts to fall,
and how predictive fit moves.

```{r plot-selection, fig.width=7, fig.height=4.5}
plot(k_selection)
```

Read these panels together rather than optimizing any one of them. A common
pattern is that coherence and exclusivity favor a moderate `k` while perplexity
keeps improving as `k` grows; the interpretable choice is usually the smallest
`k` that captures the distinct themes without splitting them into redundant
near-duplicates.

## Evaluate a Chosen Model in Depth

Once the grid points you toward a candidate, refit that single model and inspect
it more closely. `evaluate_topic_model()` returns the same metrics as the grid,
but with `level = "all"` it also reports the per-topic values, which is where you
can spot a single incoherent or non-exclusive topic that an aggregate number
would hide.

```{r evaluate}
fit <- fit_topic_model(
  dfmat,
  engine  = "topicmodels",
  model   = "lda",
  method  = "Gibbs",
  k       = 6,
  control = list(fit = list(seed = 42L, iter = 400L, burnin = 100L, thin = 1L))
)

evaluation <- evaluate_topic_model(
  fit,
  training = dfmat,
  metrics  = c("coherence_umass", "diversity", "exclusivity"),
  top_n    = 10L,
  level    = "all"
)

evaluation
```

It also helps to read the topics themselves. `get_top_terms()` turns the chosen
model into a compact, human-readable view that you can sanity-check against the
metrics.

```{r top-terms}
get_top_terms(fit, n = 8, format = "wide")
```

## Assess Stability Across Seeds

A model that changes substantially when you change the random seed is hard to
defend as a stable research object. `assess_topic_stability()` makes that check
explicit: it refits the same specification across several seeds, aligns
vocabularies, matches topics across runs, and reports matched-topic cosine
similarities. Because topic labels are arbitrary across runs, the matching step
is what makes the comparison meaningful.

```{r stability}
stability <- assess_topic_stability(
  dfmat,
  engine  = "topicmodels",
  model   = "lda",
  method  = "Gibbs",
  k       = 6,
  seeds   = 1:4,
  control = list(fit = list(iter = 400L, burnin = 100L, thin = 1L))
)

stability
```

Higher matched similarities mean the topic structure is reproducible across
seeds. If stability is low at your preferred `k`, that is useful evidence: it may
argue for a smaller `k`, more iterations, or a more constrained model before the
topics are treated as fixed findings.

## A Statistical Test for K: OpTop

The metrics and stability checks above are diagnostic, but they still leave the
final call to judgment. For a formal decision rule, the **OpTop** package
implements the chi-square optimal-topic test of Lewis and Grossetti (2022). The
test compares fitted LDA models against the document-level word distribution and
returns a statistical criterion for the optimal number of topics, rather than a
metric to eyeball.

NLPstudio does not reimplement the test. Instead it prepares the exact objects
OpTop expects and folds the result back into the same selection report. The
bridge is intentionally narrow: OpTop is written for `LDA_VEM` objects from
`topicmodels::LDA()`, so `as_optop_input()` requires a VEM grid fit with
`return_fits = TRUE` and rejects other model classes.

```{r optop-bridge, eval=FALSE}
selection_vem <- select_k_topics(
  dfmat,
  engine      = "topicmodels",
  model       = "lda",
  method      = "VEM",
  k_grid      = 4:8,
  metrics     = c("diversity", "exclusivity"),
  holdout     = 0,
  return_fits = TRUE,
  control     = list(fit = list(seed = 42L))
)

optop_input <- as_optop_input(
  selection_vem,
  weighted_dfm = as_optop_weighted_dfm(dfmat)
)

optop_result <- OpTop::optimal_topic(
  lda_models   = optop_input$lda_models,
  weighted_dfm = optop_input$weighted_dfm,
  q     = 0.8,
  alpha = 0.05,
  do_plot = FALSE
)
```

Passing the OpTop output back to `summarize_k_selection()` merges its columns
into the selection report by `k`, so the statistical test and the
interpretability metrics sit in one export-ready table.

```{r optop-summary, eval=FALSE}
k_report_optop <- summarize_k_selection(selection_vem, optop = optop_result)
data.table::fwrite(k_report_optop, tempfile(fileext = ".csv"))
```

## Choosing a Final K

There is no formula that turns this evidence into a single answer, but a
defensible workflow looks like this:

1. Use `select_k_topics()` to scan a grid and `plot()` the aggregate metrics to
   find the region where coherence and exclusivity are strong and diversity has
   not yet collapsed.
2. Refit the leading candidates and read their per-topic metrics and top terms
   with `evaluate_topic_model(level = "all")` and `get_top_terms()`.
3. Confirm with `assess_topic_stability()` that the structure survives a change
   of seed.
4. Where a formal criterion is needed, run the OpTop chi-square test and report
   it alongside the metrics via `summarize_k_selection()`.

The goal is convergent evidence: a `k` that is coherent, exclusive, non-redundant,
stable, and -- where you use it -- consistent with the statistical test.

## API Map

| Task | Primary function |
|---|---|
| Fit a grid of candidate models | `select_k_topics()` |
| Report selection evidence | `summarize_k_selection()` |
| Plot metrics against K | `plot()` on the selection object |
| Evaluate a chosen model | `evaluate_topic_model()` |
| Read topic top terms | `get_top_terms()` |
| Assess seed stability | `assess_topic_stability()` |
| Prepare OpTop inputs | `as_optop_weighted_dfm()`, `as_optop_input()` |

## References

Aletras, N., & Stevenson, M. (2013). Evaluating topic coherence using
distributional semantics. _Proceedings of the 14th Conference of the European
Chapter of the Association for Computational Linguistics_, 13-22.

Lewis, C. M., & Grossetti, F. (2022).
[A statistical approach for optimal topic model identification](https://www.jmlr.org/papers/volume23/19-297/19-297.pdf).
_Journal of Machine Learning Research_, 23(58), 1-20.

Mimno, D., Wallach, H., Talley, E., Leenders, M., & McCallum, A. (2011).
Optimizing semantic coherence in topic models. _Proceedings of the 2011
Conference on Empirical Methods in Natural Language Processing_, 262-272.
