---
title: "Corpus Preparation and Text Analysis"
subtitle: "From SEC-Style JSON to Quanteda Workflows"
author: "by Francesco Grossetti"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Corpus Preparation and Text Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
has_json <- requireNamespace("RcppSimdJson", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_json
)
```

NLPstudio is designed for workflows where text starts outside R: scraped
filings, structured JSON records, large corpora, and document metadata that
needs to stay aligned with the text. The topic-model vignette starts after the
modeling input is already prepared. This vignette covers the earlier part of
the workflow: getting filing text into a reliable corpus and producing common
document-level features.

The examples use five bundled 10-K JSON filings. They are small enough to ship
with the package but realistic enough to show the intended data path. Each JSON
file contains filing metadata and top-level `item_*` fields, such as `item_1`,
`item_1A`, and `item_7`.

The JSON import chunks require the optional **RcppSimdJson** package. If it is
not installed, the vignette still knits and shows the code, but the JSON-backed
chunks are not evaluated.

## Locate the Example JSON Files

The example files live in `inst/extdata/json/` in the source package. Once the
package is installed, they are available through `system.file()`.

```{r locate-json}
library(NLPstudio)
library(data.table)
library(quanteda)

json_dir <- system.file("extdata", "json", package = "NLPstudio")
if (!nzchar(json_dir)) {
  json_dir <- file.path("..", "inst", "extdata", "json")
}

json_files <- list.files(
  json_dir,
  pattern = "_10K_.*\\.json$",
  full.names = TRUE
)

basename(json_files)
```

## Flatten JSON to a Filing-Item Table

`from_json_to_df()` reads the JSON files, identifies the 10-K item fields, and
returns one row per filing item. Metadata such as CIK, company name, filing
type, filing date, period of report, SIC, and filename stay attached to each
item row.

```{r import-json}
filings <- from_json_to_df(
  json_files,
  what = "10-K",
  ncores = 1,
  max_chunk_size = 1
)

nrow(filings)
filings[, .N, by = .(company, filing_type, fyear)]
filings[, .N, by = item][order(item)]
```

The full imported table contains all non-empty 10-K sections from the five
filings. For the rest of the vignette, we focus on three sections that are easy
to interpret: business description (`item_1`), risk factors (`item_1A`), and
management discussion (`item_7`).

```{r focus-items}
analysis_dt <- filings[
  item %in% c("item_1", "item_1A", "item_7"),
  .(cik, company, filing_type, filing_date, period_of_report,
    fyear, sic, filename, item, text)
]

analysis_dt[, text_chars := nchar(text)]
analysis_dt[, .(company, item, text_chars)][order(company, item)]
```

## Build a Quanteda Corpus

`define_corpus()` turns the filing-item table into a `quanteda` corpus. It uses
the filing filename and item name to create stable document identifiers, while
the remaining columns become document variables.

```{r define-corpus}
corp <- define_corpus(data.table::copy(analysis_dt))

summary(corp, n = 6)
quanteda::docvars(corp)[1:6, c("company", "item", "fyear")]
```

At this point, each filing item is a separate document. This is usually the most
useful unit for downstream analysis because risk factors, business descriptions,
and management discussion sections often answer different research questions.

## Tokenize and Build a DFM

`tokenize_corpus()` wraps `quanteda::tokens()` and keeps document order stable.
For small examples, a single core is enough. On larger corpora, the same
function can split the corpus into chunks and process them through the package's
parallel helper.

```{r tokenize}
toks <- tokenize_corpus(
  corp,
  ncores = 1,
  remove_punct = TRUE,
  remove_numbers = TRUE,
  remove_symbols = TRUE
)

toks <- quanteda::tokens_tolower(toks)
toks <- quanteda::tokens_remove(toks, pattern = quanteda::stopwords("en"))

dfmat <- quanteda::dfm(toks)
dfmat <- quanteda::dfm_trim(dfmat, min_termfreq = 5)

dfmat
quanteda::topfeatures(dfmat, 15)
```

Singularization is optional, but it is often helpful when the analysis should
count plural and singular nouns together. `singularize_tokens()` applies
NLPstudio's internal English singularization rules to the token vocabulary and
then replaces only the affected token types, so the operation remains fast even
when the corpus is much larger than this example.

```{r singularize}
singular_toks <- singularize_tokens(toks, ncores = 1, min_char = 4)
singular_dfm <- quanteda::dfm(singular_toks)
quanteda::topfeatures(singular_dfm, 15)
```

## Apply a Dictionary

NLPstudio ships domain dictionaries as `quanteda` dictionary objects. Dictionary
lookup is useful when the construct is known in advance and the goal is to
measure document-level exposure to a curated vocabulary.

```{r dictionary-lookup}
fls_toks <- lookup_tokens(
  toks,
  dictionary = data_dictionary_BozanicRoulstoneVanBuskirk_FLS,
  ncores = 1,
  exclusive = FALSE
)

fls_dfm <- quanteda::dfm(fls_toks)
quanteda::topfeatures(fls_dfm, 10)
```

The same pattern works with the other bundled dictionaries, including firm
complexity, corporate social responsibility, sustainable development goals, and
the Feng Li forward-looking statement dictionary.

## Preprocessing the Tokens

Beyond singularization, NLPstudio wraps the most common token-preprocessing
steps so the workflow stays in one place and keeps the same parallel controls
(`ncores`, `nchunks`, `socket`).

`ngram_tokens()` builds n-grams (and skip-grams) from a tokens object.

```{r ngrams}
bigrams <- ngram_tokens(toks, n = 2, ncores = 1)
quanteda::topfeatures(quanteda::dfm(bigrams), 8)
```

Rather than fixing phrases by hand, `detect_collocations()` scores candidate
multiword expressions, and `compound_tokens()` binds the ones you keep into
single tokens. The two functions compose directly.

```{r collocations}
colls <- detect_collocations(toks, size = 2, min_count = 3)
head(colls, 5)

toks_comp <- compound_tokens(toks, pattern = head(colls, 10), ncores = 1)
```

`stem_tokens()` applies the Snowball stemmer (many languages), reducing related
word forms to a common stem.

```{r stem}
stemmed <- stem_tokens(toks, ncores = 1)
quanteda::topfeatures(quanteda::dfm(stemmed), 8)
```

`lemmatize_tokens()` maps tokens to lemmas. The dependency-free `"lookup"` engine
applies a user-supplied lemma map; the optional `"spacy"` engine derives lemmas
with **spacyr** (see [parse_corpus()]).

```{r lemmatize-lookup}
lemma_map <- data.frame(
  token = c("companies", "risks", "costs", "controls"),
  lemma = c("company", "risk", "cost", "control")
)
lemmatized <- lemmatize_tokens(toks, engine = "lookup", lemma = lemma_map)
quanteda::topfeatures(quanteda::dfm(lemmatized), 8)
```

```{r lemmatize-spacy, eval=FALSE}
# Requires a working spaCy installation; see parse_corpus().
spacyr::spacy_initialize(model = "en_core_web_sm")
lemmatize_tokens(toks, engine = "spacy")
```

Finally, `weight_dfm()` exposes TF-IDF and the other quanteda weighting schemes
on a dfm.

```{r weight}
tfidf <- weight_dfm(dfmat, scheme = "tfidf")
tfidf
```

## Reshape and Summarize the Corpus

`reshape_corpus()` changes the document unit. For example, item-level documents
can be split into sentence-level documents before computing readability or
other document-level statistics.

```{r reshape}
sentence_corp <- reshape_corpus(corp, to = "sentences", ncores = 1)

quanteda::ndoc(corp)
quanteda::ndoc(sentence_corp)
```

`summarize_corpus()` returns a compact table of document-level corpus statistics
from `quanteda.textstats`.

```{r summarize-corpus}
corpus_summary <- summarize_corpus(corp, ncores = 1)

corpus_summary[1:6]
```

## Readability

Readability scores are often useful when filing sections differ substantially in
length and complexity. `calculate_readability()` keeps the same corpus input and
returns a `data.table`.

```{r readability}
readability <- calculate_readability(
  corp,
  ncores = 1,
  measure = c("Flesch", "Flesch.Kincaid")
)

readability[1:6]
```

## Similarity and Distance

`calculate_similarity()` and `calculate_distance()` work on a `quanteda` dfm.
They expose quanteda's internal threaded computation through `ncores`, avoiding
the incorrect pattern of splitting an all-pairs matrix across document chunks.

```{r similarity-distance}
similarity <- calculate_similarity(
  dfmat,
  ncores = 1,
  margin = "documents",
  method = "cosine"
)

distance <- calculate_distance(
  dfmat,
  ncores = 1,
  margin = "documents",
  method = "euclidean"
)

round(as.matrix(similarity)[1:5, 1:5], 3)
round(as.matrix(distance)[1:5, 1:5], 3)
```

## Export Tables

Most outputs are ordinary tabular objects or can be converted to tabular form.
That makes it straightforward to save intermediate results for review,
appendix tables, or downstream modeling.

```{r export}
export_readability <- data.table::copy(readability)
export_readability[, source := "bundled_10k_json"]

out_file <- tempfile(fileext = ".csv")
data.table::fwrite(export_readability, out_file)
out_file
```

## Optional spaCy Parsing

`parse_corpus()` wraps **spacyr** for part-of-speech parsing and related
annotations. It requires a working Python spaCy installation, so it is shown
but not evaluated in this vignette.

```{r spacy-parse, eval=FALSE}
spacyr::spacy_initialize(model = "en_core_web_sm")

parsed <- parse_corpus(
  corp,
  ncores = 1,
  lemma = TRUE,
  pos = TRUE
)

head(parsed)
```

## Workflow Map

The non-topic-model API is designed to be composable:

| Task | Primary function |
|---|---|
| Flatten SEC-style JSON | `from_json_to_df()` |
| Build a quanteda corpus | `define_corpus()` |
| Tokenize text | `tokenize_corpus()` |
| Singularize tokens | `singularize_tokens()` |
| Apply dictionaries | `lookup_tokens()` |
| Build n-grams | `ngram_tokens()` |
| Detect collocations | `detect_collocations()` |
| Compound phrases | `compound_tokens()` |
| Stem tokens | `stem_tokens()` |
| Lemmatize tokens | `lemmatize_tokens()` |
| Weight a dfm | `weight_dfm()` |
| Reshape documents | `reshape_corpus()` |
| Summarize corpus | `summarize_corpus()` |
| Compute readability | `calculate_readability()` |
| Compute similarity | `calculate_similarity()` |
| Compute distance | `calculate_distance()` |
| Optional spaCy parsing | `parse_corpus()` |
