---
title: "Topic Model Selection and Diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Topic Model Selection and Diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
has_topicmodels <- requireNamespace("topicmodels", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_topicmodels
)
```

This vignette shows the topic-model workflow introduced across the `v0.8.x` and
`v0.9.0` releases: fit one model, evaluate it, compare candidate values of K,
assess repeated-fit stability across seeds, summarize topics, and export the
interpretation table.

```{r data}
library(NLPstudio)
library(quanteda)

docs <- data.frame(
  doc_id = paste0("doc", 1:8),
  text = c(
    "Revenue growth improved after subscription demand increased.",
    "Recurring software revenue supported customer retention.",
    "Operating margin expanded as cloud infrastructure costs declined.",
    "Capital allocation reflected debt covenants and interest expense.",
    "Audit committee oversight focused on internal control remediation.",
    "Risk disclosures emphasized liquidity and refinancing pressure.",
    "Cybersecurity controls shaped technology governance disclosures.",
    "Management discussed compliance monitoring and reporting quality."
  ),
  group = rep(c("performance", "governance"), each = 4),
  stringsAsFactors = FALSE
)

corp <- quanteda::corpus(docs, text_field = "text", docid_field = "doc_id")
toks <- quanteda::tokens(corp, remove_punct = TRUE)
toks <- quanteda::tokens_tolower(toks)
toks <- quanteda::tokens_remove(toks, pattern = quanteda::stopwords("en"))
dfm <- quanteda::dfm(toks)
```

Fit one model first. `fit_topic_model()` remains the primitive: one call
estimates one model specification.

```{r fit}
fit <- fit_topic_model(
  dfm,
  engine = "topicmodels",
  model = "lda",
  method = "Gibbs",
  k = 2,
  doc_data = docs,
  control = list(fit = list(seed = 1L, iter = 75L, burnin = 0L, thin = 1L))
)

get_top_terms(fit, n = 5)
```

Use `evaluate_topic_model()` when the topic count is already fixed and you want
quality metrics for a fitted model.

```{r evaluate}
evaluate_topic_model(
  fit,
  training = dfm,
  metrics = c("coherence_umass", "diversity", "exclusivity"),
  top_n = 5L,
  level = "all"
)
```

Use `select_k_topics()` when candidate values of K should be compared under the
same model family and controls. Existing defaults are unchanged; stability is
added only when `stability_seeds` is supplied.

```{r select-k}
k_selection <- select_k_topics(
  dfm,
  engine = "topicmodels",
  model = "lda",
  method = "Gibbs",
  k_grid = 2:3,
  metrics = c("coherence_umass", "diversity", "exclusivity"),
  top_n = 5L,
  holdout = 0,
  seed = 10L,
  stability_seeds = c(1L, 2L),
  control = list(fit = list(iter = 75L, burnin = 0L, thin = 1L))
)

k_selection
attr(k_selection, "stability")$k2
```

`assess_topic_stability()` is a transparent repeated-fit wrapper. It calls
`fit_topic_model()` with the same data, engine, model, K, method, controls, and
extra arguments, changing only the seed unless `resampling` is supplied.

```{r stability-auto}
stability <- assess_topic_stability(
  dfm,
  engine = "topicmodels",
  model = "lda",
  method = "Gibbs",
  k = 2,
  seeds = 1:3,
  control = list(fit = list(iter = 75L, burnin = 0L, thin = 1L))
)

stability
```

Users who prefer to fit models manually can pass a list of `nlp_topic_fit`
objects and still use package-owned topic matching and scoring.

```{r stability-manual}
fits <- lapply(1:3, function(seed) {
  fit_topic_model(
    dfm,
    engine = "topicmodels",
    model = "lda",
    method = "Gibbs",
    k = 2,
    control = list(fit = list(seed = seed, iter = 75L, burnin = 0L, thin = 1L))
  )
})

assess_topic_stability(fits, seeds = 1:3)
```

Finally, `summarize_topics()` builds a compact interpretation table with top
terms, prevalence, available metrics, and representative documents.

```{r summarize}
topic_summary <- summarize_topics(
  fit,
  training = dfm,
  doc_data = docs,
  top_n = 5L,
  representative_n = 2L,
  include_text = TRUE
)

topic_summary
```

The summary includes a nested `representative_documents` column for metadata.
For flat-file export, drop that list column and keep the collapsed document IDs
and text fields.

```{r export}
export_summary <- data.table::copy(topic_summary)
export_summary[, representative_documents := NULL]
data.table::fwrite(export_summary, tempfile(fileext = ".csv"))
```

