---
title: "Topic Model API and Usage"
subtitle: "A Consistent Workflow Across Topic-Model Backends"
author: "by Francesco Grossetti"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Topic Model API and Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
has_topicmodels <- requireNamespace("topicmodels", quietly = TRUE)
has_tidytext <- requireNamespace("tidytext", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_topicmodels
)
```

Topic models are useful, but the R ecosystem exposes them through several
different object types, argument names, and output formats. That makes it harder
than it should be to compare models, reuse diagnostics, or move a workflow from
one backend to another.

NLPstudio's topic-model API is built to remove that friction. The package gives
you one workflow for fitting, inspecting, predicting, evaluating, selecting, and
summarizing topic models. The backend can change, but the main verbs and the
standardized outputs stay the same.

This vignette walks through that workflow from start to finish. We will prepare
a small corpus, fit one topic model, inspect the standardized outputs, evaluate
the model, compare candidate values of `k`, assess seed stability, build an
interpretation table, and sketch how the same calls transfer to other engines.

The notation follows Lewis and Grossetti (2022). In particular, this vignette
uses DTW for document-topic weights and TWW for topic-word weights. Those two
tables are the main package-standard representation of a fitted topic model:
DTW describes documents in the topic space, while TWW describes topics in the
vocabulary space.

The package currently supports topic models from **text2vec**, **topicmodels**,
**seededlda**, and optionally **topicmodels.etm**. To keep this vignette fast and
reliable during package checks, only the **topicmodels** examples are evaluated.
The other backend sections are included as non-evaluated templates so you can see
the same API shape without requiring every optional backend on your machine.

## Prepare Data Once

The workflow starts with ordinary data: one row per document, one text column,
and any metadata that helps interpret the documents later. Here the metadata is
minimal, but the same pattern works with richer document-level fields such as
firm identifiers, filing dates, industries, countries, or human-coded labels.

Topic models are estimated from a document-feature matrix. The metadata does not
enter the model directly in this example, but we keep it next to the model so it
can be returned by downstream helpers. That is what lets later calls attach text,
groups, and document identifiers to predictions, representative documents, and
summary tables.

```{r data}
library(NLPstudio)
library(quanteda)

docs <- data.frame(
  doc_id = paste0("doc", 1:10),
  text = c(
    "Revenue growth improved after subscription demand increased.",
    "Recurring software revenue supported customer retention.",
    "Cloud infrastructure costs declined and operating margin expanded.",
    "Capital allocation reflected debt covenants and interest expense.",
    "Audit committee oversight focused on internal control remediation.",
    "Risk disclosures emphasized liquidity and refinancing pressure.",
    "Cybersecurity controls shaped technology governance disclosures.",
    "Management discussed compliance monitoring and reporting quality.",
    "Subscription renewals improved annual recurring revenue trends.",
    "Debt refinancing reduced interest expense and liquidity risk."
  ),
  group = rep(c("performance", "governance"), each = 5),
  stringsAsFactors = FALSE
)

corp <- quanteda::corpus(docs, text_field = "text", docid_field = "doc_id")
toks <- quanteda::tokens(corp, remove_punct = TRUE)
toks <- quanteda::tokens_tolower(toks)
toks <- quanteda::tokens_remove(toks, pattern = quanteda::stopwords("en"))
dfm <- quanteda::dfm(toks)
dfm
```

## Fit One Model

`fit_topic_model()` is the basic modeling verb. One call means one model
specification: one input matrix, one engine, one model family, one value of `k`,
and one set of backend controls.

The object returned by `fit_topic_model()` is an `nlp_topic_fit`. It keeps the
original backend model, but it also stores the package-owned pieces that make the
rest of the API consistent: document-topic weights (DTW), topic-word weights
(TWW), stable topic identifiers, the fitted vocabulary, optional document
metadata, backend controls, and available hyperparameters.

```{r fit}
fit <- fit_topic_model(
  dfm,
  engine = "topicmodels",
  model = "lda",
  method = "Gibbs",
  k = 2,
  doc_data = docs,
  control = list(fit = list(seed = 1L, iter = 75L, burnin = 0L, thin = 1L))
)

fit
```

The printed object is intentionally compact. It tells you what was fit and gives
you enough context to confirm that the model has the expected engine, model
family, topic count, document count, and vocabulary size.

Notice the topic labels. Standardized outputs use identifiers such as
`Topic001` and `Topic002`. Those labels are generated by NLPstudio rather than
by a specific backend, which is what allows the same helper functions to work
across engines.

## Inspect Standardized Outputs

After fitting, the first question is usually: "What did the model learn?"
NLPstudio answers that with two standardized tables.

Use `get_dtw()` for document-topic weights. Each row corresponds to a document,
and the `Topic###` columns describe how much of that document is associated with
each topic. Use `get_tww()` for topic-word weights. Each row corresponds to a
topic, and the columns describe the topic's distribution over the fitted
vocabulary.

```{r standardized-outputs}
dtw <- get_dtw(fit, docvars = TRUE, include_text = TRUE)
tww <- get_tww(fit)

head(dtw)
tww[, 1:6, with = FALSE]
```

The full TWW matrix is useful for computation, but it is often too wide for
human reading. When the goal is interpretation, `get_top_terms()` returns the
most important words for each topic in either long or wide format. It accepts a
fitted model or a TWW table, so you can use it early in exploration or later in
a reporting pipeline.

```{r top-terms}
top_terms <- get_top_terms(fit, n = 5, format = "long")
top_terms

get_top_terms(tww, n = 3, format = "wide")
```

Some engines expose topic-model hyperparameters, while others expose only part
of that information. NLPstudio returns what is available through
`get_topic_hyperparameters()` in a backend-neutral table. The original backend
controls are still stored on `fit$backend_control`, which is helpful when you
need to audit exactly how a model was estimated.

```{r hyperparameters}
get_topic_hyperparameters(fit)
```

## Interpret Topics

Topic interpretation usually combines two views. The first is lexical: which
words define the topic? The second is documentary: which documents are strongest
examples of the topic?

For the lexical view, `plot_top_terms()` uses the long output from
`get_top_terms()`. The plot is a quick check on whether the discovered topics
look distinct and interpretable.

```{r plot-top-terms, eval=has_topicmodels && has_tidytext, fig.width=7, fig.height=4}
plot_top_terms(top_terms)
```

`plot_dtw()` gives the document-side view. It shows how topic weights are spread
across documents, which helps you see whether the fitted topics dominate a few
documents, appear broadly across the corpus, or remain very diffuse.

```{r plot-dtw, fig.width=7, fig.height=4}
plot_dtw(fit)
```

Representative documents are often the easiest way to name and validate a
topic. `get_representative_candidates()` ranks documents by their dominant topic
weight and assigns simple candidate bands. If metadata or text are available,
the function can return them with the ranking so you can inspect the examples in
context.

```{r representatives}
representatives <- get_representative_candidates(
  fit,
  doc_data = docs,
  include_text = TRUE
)

representatives[
  ,
  .(doc_id, topic_max_id, topic_max_value, candidate_band, topic_rank, text)
]
```

## Predict New Documents

Once a model is fit, you may want to score new documents against the same topic
space. `predict_topic_model()` handles the practical detail that new documents
rarely have exactly the same vocabulary columns as the training matrix. The new
input is aligned to the fitted vocabulary before backend prediction, and the
returned DTW table uses the same `Topic###` naming contract as the fitted model.

```{r predict}
new_docs <- data.frame(
  doc_id = paste0("new", 1:3),
  text = c(
    "Recurring revenue improved after subscription renewals.",
    "Audit remediation and compliance controls remained a priority.",
    "Debt refinancing lowered liquidity risk and interest expense."
  ),
  source = c("forecast", "governance", "capital"),
  stringsAsFactors = FALSE
)

new_corp <- quanteda::corpus(new_docs, text_field = "text", docid_field = "doc_id")
new_toks <- quanteda::tokens(new_corp, remove_punct = TRUE)
new_toks <- quanteda::tokens_tolower(new_toks)
new_toks <- quanteda::tokens_remove(new_toks, pattern = quanteda::stopwords("en"))
new_dfm <- quanteda::dfm(new_toks)

pred <- predict_topic_model(
  fit,
  new_dfm,
  doc_data = new_docs,
  include_text = TRUE
)

pred
```

## Evaluate Model Quality

Interpretability is not only visual. `evaluate_topic_model()` collects common
quality metrics into one long-format table so they can be compared, plotted, or
joined to model-selection results. Some metrics summarize the whole model;
others are defined topic by topic. Setting `level = "all"` returns both where
available.

```{r evaluate}
evaluation <- evaluate_topic_model(
  fit,
  training = dfm,
  metrics = c("coherence_umass", "diversity", "exclusivity", "train_perplexity"),
  top_n = 5L,
  level = "all"
)

evaluation
```

## Select K

Choosing the number of topics is a modeling decision, not just a software
argument. `select_k_topics()` makes that decision easier to inspect by fitting a
grid of candidate topic counts and returning the requested metrics in one
selection table.

The default behavior remains the simple path: fit each value of `k` once and
evaluate it. When `stability_seeds` is supplied, each candidate `k` is also
refit across those seeds. The selection table then includes aggregate stability,
and the full stability details are attached as an attribute for deeper
inspection.

```{r select-k}
k_selection <- select_k_topics(
  dfm,
  engine = "topicmodels",
  model = "lda",
  method = "Gibbs",
  k_grid = 2:3,
  metrics = c("coherence_umass", "diversity", "exclusivity"),
  top_n = 5L,
  holdout = 0,
  seed = 10L,
  stability_seeds = c(1L, 2L),
  control = list(fit = list(iter = 75L, burnin = 0L, thin = 1L))
)

k_selection
attr(k_selection, "stability")$k2
```

## Assess Stability Directly

Topic models can move when the random seed changes. That does not necessarily
mean a model is bad, but it is information you should see before treating topic
labels as stable research objects.

`assess_topic_stability()` makes that check explicit. In the automatic workflow,
it repeatedly calls `fit_topic_model()` with the same model specification and
different seeds. Unless you supply `resampling`, the seed is the only thing that
changes across runs.

Topic labels are arbitrary across repeated runs. A topic called `Topic001` in
one seed may correspond to `Topic002` in another seed. The stability helper
therefore extracts standardized TWW from each run, aligns vocabularies, matches
topics across runs, and reports matched-topic cosine similarities.

```{r stability-auto}
stability <- assess_topic_stability(
  dfm,
  engine = "topicmodels",
  model = "lda",
  method = "Gibbs",
  k = 2,
  seeds = 1:3,
  control = list(fit = list(iter = 75L, burnin = 0L, thin = 1L))
)

stability
```

If you prefer to fit models yourself, you can still use the same matching and
scoring code. Pass a list of `nlp_topic_fit` objects, and
`assess_topic_stability()` will skip fitting and move directly to standardized
TWW extraction, topic matching, and stability scoring.

```{r stability-manual}
fits <- lapply(1:3, function(seed) {
  fit_topic_model(
    dfm,
    engine = "topicmodels",
    model = "lda",
    method = "Gibbs",
    k = 2,
    control = list(fit = list(seed = seed, iter = 75L, burnin = 0L, thin = 1L))
  )
})

assess_topic_stability(fits, seeds = 1:3)
```

## Summarize and Export

After fitting, inspecting, evaluating, and checking stability, the next step is
usually a table that can be reviewed, exported, or adapted for a paper appendix.
`summarize_topics()` builds that table directly from the fitted object. It
returns one row per topic with top terms, top-term probabilities, topic
prevalence, available metrics, representative documents, and optional text or
metadata.

```{r summarize}
topic_summary <- summarize_topics(
  fit,
  training = dfm,
  doc_data = docs,
  top_n = 5L,
  representative_n = 2L,
  include_text = TRUE
)

topic_summary
```

The summary includes a `representative_documents` list column. That is useful
inside R because each topic can carry its own small document table. For flat file
exports, drop the list column and write the remaining scalar columns.

```{r export}
export_summary <- data.table::copy(topic_summary)
export_summary[, representative_documents := NULL]
data.table::fwrite(export_summary, tempfile(fileext = ".csv"))
```

## Migrate Legacy WarpLDA Outputs

Earlier versions of NLPstudio exposed `warp_lda()` as a direct wrapper around
the **text2vec** WarpLDA estimator. That fitting function is no longer part of
the public API, but saved outputs from old projects can still be brought into
the current workflow. `as_nlp_topic_fit()` converts the old list structure into
an `nlp_topic_fit` object without refitting the model.

Legacy objects usually contain `lda_object`, `theta`, and `phi`. The adapter
uses `theta` as DTW, `phi` as TWW, and keeps the original WarpLDA backend object
as `model_object`.

```{r legacy-warplda, eval=FALSE}
old <- readRDS("legacy-warp-lda-output.rds")
legacy_fit <- as_nlp_topic_fit(old)

get_dtw(legacy_fit)
get_tww(legacy_fit)
get_top_terms(legacy_fit, n = 10)
summarize_topics(legacy_fit, training = original_dfm)
```

If the old object was saved without `theta` or `phi`, the adapter preserves the
parts that are available and warns about the missing cache. Functions that need
the missing component will then fail with the same messages used by current
partial `nlp_topic_fit` objects.

## Backend Portability

The point of the API is that the workflow above does not belong only to one
backend. The fitting controls differ by engine, but the surrounding calls stay
recognizable: fit a model, extract DTW/TWW, interpret top terms, evaluate, and
summarize.

The examples below are not evaluated in this vignette because optional backend
availability varies across machines and CI environments. They are meant as
templates for moving the same workflow to another estimator.

```{r text2vec-backend, eval=FALSE}
fit_text2vec <- fit_topic_model(
  dfm,
  engine = "text2vec",
  model = "lda",
  k = 2,
  control = list(fit = list(n_iter = 100, progressbar = FALSE))
)

get_top_terms(fit_text2vec, n = 5)
evaluate_topic_model(fit_text2vec, training = dfm, metrics = c("diversity", "exclusivity"))
```

```{r seededlda-backend, eval=FALSE}
fit_seededlda <- fit_topic_model(
  dfm,
  engine = "seededlda",
  model = "lda",
  k = 2
)

get_dtw(fit_seededlda)
get_tww(fit_seededlda)
```

Seeded LDA uses the same wrapper but requires a dictionary. Once the model is
fit, the downstream accessors remain the same.

```{r seededlda-dictionary, eval=FALSE}
seed_dict <- quanteda::dictionary(list(
  performance = c("revenue", "subscription", "margin"),
  governance = c("audit", "control", "compliance")
))

fit_seeded <- fit_topic_model(
  dfm,
  engine = "seededlda",
  model = "seededlda",
  dictionary = seed_dict
)

summarize_topics(fit_seeded, doc_data = docs)
```

## Optional ETM Extensions

Embedded topic models are an optional advanced path. They are available when
**topicmodels.etm**, **torch**, and a working torch backend are installed. ETM
fits still use the same downstream accessors for DTW, TWW, top terms,
prediction, evaluation, and summaries. They also expose ETM-specific topic and
term embeddings when those are present on the fitted object.

```{r etm-backend, eval=FALSE}
fit_etm <- fit_topic_model(
  dfm,
  engine = "topicmodels.etm",
  model = "etm",
  k = 2,
  control = list(
    model = list(embeddings = 50L),
    fit = list(epochs = 10L, batch_size = 8L)
  )
)

get_topic_embeddings(fit_etm)
get_term_embeddings(fit_etm)
plot_topic_embeddings(fit_etm, top_n = 10)
```

## API Map

The main topic-model API is intentionally layered. You can use only the pieces
you need, but the functions are designed to compose into one continuous
workflow:

| Task | Primary function |
|---|---|
| Fit one model | `fit_topic_model()` |
| Extract DTW/TWW | `get_dtw()`, `get_tww()` |
| Interpret terms and documents | `get_top_terms()`, `get_representative_candidates()` |
| Plot standardized outputs | `plot_top_terms()`, `plot_dtw()` |
| Predict new documents | `predict_topic_model()` |
| Evaluate quality | `evaluate_topic_model()` |
| Select K | `select_k_topics()` |
| Assess seed stability | `assess_topic_stability()` |
| Build interpretation tables | `summarize_topics()` |
| Migrate legacy WarpLDA output | `as_nlp_topic_fit()` |
| Use ETM embeddings | `get_topic_embeddings()`, `get_term_embeddings()`, `plot_topic_embeddings()` |

## References

Aletras, N., & Stevenson, M. (2013). Evaluating topic coherence using
distributional semantics. _Proceedings of the 14th Conference of the European
Chapter of the Association for Computational Linguistics_, 13-22.

Blei, D. M., Ng, A. Y., & Jordan, M. I. (2003).
[Latent Dirichlet Allocation](https://www.jmlr.org/papers/volume3/blei03a/blei03a.pdf).
_Journal of Machine Learning Research_, 3, 993-1022.

Blei, D. M., & Lafferty, J. D. (2006).
[Correlated topic models](https://papers.nips.cc/paper/2906-correlated-topic-models).
_Advances in Neural Information Processing Systems_, 18, 147.

Blei, D. M., & Lafferty, J. D. (2007).
[A correlated topic model of Science](https://doi.org/10.1214/07-AOAS114).
_The Annals of Applied Statistics_, 1(1), 17-35.

Chen, J., Li, K., Zhu, J., & Chen, W. (2016).
[WarpLDA: A Cache Efficient O(1) Algorithm for Latent Dirichlet Allocation](https://pacman.cs.tsinghua.edu.cn/~cwg/publication/vldb16/vldb16.pdf).
_Proceedings of the VLDB Endowment_, 9(10), 744-755.

Dieng, A. B., Ruiz, F. J. R., & Blei, D. M. (2020).
[Topic Modeling in Embedding Spaces](https://arxiv.org/pdf/1907.04907).
_Transactions of the Association for Computational Linguistics_, 8, 439-453.

Du, L., Buntine, W. L., Jin, H., & Chen, C. (2012).
[Sequential latent Dirichlet allocation](https://doi.org/10.1007/s10115-011-0425-1).
_Knowledge and Information Systems_, 31(3), 475-503.

Jagarlamudi, J., Daume III, H., & Udupa, R. (2012).
[Incorporating lexical priors into topic models](https://aclanthology.org/E12-1021.pdf).
In _Proceedings of the 13th Conference of the European Chapter of the
Association for Computational Linguistics_, 204-213.

Lewis, C. M., & Grossetti, F. (2022).
[A statistical approach for optimal topic model identification](https://www.jmlr.org/papers/volume23/19-297/19-297.pdf).
_Journal of Machine Learning Research_, 23(58), 1-20.

Lu, B., Ott, M., Cardie, C., & Tsou, B. K. (2011).
[Multi-aspect sentiment analysis with topic models](https://doi.org/10.1109/ICDMW.2011.125).
In _2011 IEEE 11th International Conference on Data Mining Workshops_, 81-88.

Mimno, D., Wallach, H., Talley, E., Leenders, M., & McCallum, A. (2011).
Optimizing semantic coherence in topic models. _Proceedings of the 2011
Conference on Empirical Methods in Natural Language Processing_, 262-272.

Roberts, M. E., Stewart, B. M., Tingley, D., Lucas, C., Leder-Luis, J.,
Gadarian, S. K., Albertson, B., & Rand, D. G. (2014). Structural topic models
for open-ended survey responses. _American Journal of Political Science_,
58(4), 1064-1082.

Watanabe, K., & Zhou, Y. (2022).
Theory-Driven Analysis of Large Corpora: Semisupervised Topic Classification of
the UN Speeches. _Social Science Computer Review_, 40(2), 346-366.
https://doi.org/10.1177/0894439320907027.

Watanabe, K., & Baturo, A. (2024).
Seeded Sequential LDA: A Semi-Supervised Algorithm for Topic-Specific Analysis
of Sentences. _Social Science Computer Review_, 42(1), 224-248.
https://doi.org/10.1177/08944393231178605.
