---
title: "Topic Model API and Usage"
subtitle: "A Consistent Workflow Across Topic-Model Backends"
author: "by Francesco Grossetti"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Topic Model API and Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
has_topicmodels <- requireNamespace("topicmodels", quietly = TRUE)
has_tidytext <- requireNamespace("tidytext", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = has_topicmodels
)
```

NLPstudio's topic-model API is designed around one core idea: fit topic models
through a consistent interface, then work with standardized outputs regardless
of the backend that estimated the model.

The API supports topic models from **text2vec**, **topicmodels**,
**seededlda**, and optionally **topicmodels.etm**. The examples below use
**topicmodels** as the only evaluated backend so the vignette remains reliable
during package checks. Later sections show the equivalent call shapes for the
other backends without executing them.

## Prepare Data Once

Start from a regular in-memory document table. The model input is a
document-feature matrix, and the metadata table can be stored on the fit object
for downstream extraction, representative-document selection, prediction, and
topic summaries.

```{r data}
library(NLPstudio)
library(quanteda)

docs <- data.frame(
  doc_id = paste0("doc", 1:10),
  text = c(
    "Revenue growth improved after subscription demand increased.",
    "Recurring software revenue supported customer retention.",
    "Cloud infrastructure costs declined and operating margin expanded.",
    "Capital allocation reflected debt covenants and interest expense.",
    "Audit committee oversight focused on internal control remediation.",
    "Risk disclosures emphasized liquidity and refinancing pressure.",
    "Cybersecurity controls shaped technology governance disclosures.",
    "Management discussed compliance monitoring and reporting quality.",
    "Subscription renewals improved annual recurring revenue trends.",
    "Debt refinancing reduced interest expense and liquidity risk."
  ),
  group = rep(c("performance", "governance"), each = 5),
  stringsAsFactors = FALSE
)

corp <- quanteda::corpus(docs, text_field = "text", docid_field = "doc_id")
toks <- quanteda::tokens(corp, remove_punct = TRUE)
toks <- quanteda::tokens_tolower(toks)
toks <- quanteda::tokens_remove(toks, pattern = quanteda::stopwords("en"))
dfm <- quanteda::dfm(toks)
dfm
```

## Fit One Model

`fit_topic_model()` is the one-model primitive. One call estimates one model
specification. The returned `nlp_topic_fit` keeps the backend model object, but
also stores standardized document-topic weights (DTW), topic-word weights
(TWW), topic identifiers, vocabulary, document metadata, backend controls, and
hyperparameters.

```{r fit}
fit <- fit_topic_model(
  dfm,
  engine = "topicmodels",
  model = "lda",
  method = "Gibbs",
  k = 2,
  doc_data = docs,
  control = list(fit = list(seed = 1L, iter = 75L, burnin = 0L, thin = 1L))
)

fit
```

All standardized topic outputs use stable topic identifiers such as `Topic001`
and `Topic002`. That convention is the contract that makes downstream helpers
backend-agnostic.

## Inspect Standardized Outputs

Use `get_dtw()` for document-topic weights and `get_tww()` for topic-word
weights. Both return tabular outputs with package-owned naming.

```{r standardized-outputs}
dtw <- get_dtw(fit, docvars = TRUE, include_text = TRUE)
tww <- get_tww(fit)

head(dtw)
tww[, 1:6, with = FALSE]
```

Use `get_top_terms()` when the goal is interpretation rather than a full TWW
matrix. It accepts fitted models or already-standardized TWW tables.

```{r top-terms}
top_terms <- get_top_terms(fit, n = 5, format = "long")
top_terms

get_top_terms(tww, n = 3, format = "wide")
```

Hyperparameters are also exposed through a backend-neutral table. The original
backend controls remain available on `fit$backend_control` for auditability.

```{r hyperparameters}
get_topic_hyperparameters(fit)
```

## Interpret Topics

The same standardized outputs support plotting and representative-document
selection. `plot_top_terms()` uses the long output from `get_top_terms()`.

```{r plot-top-terms, eval=has_topicmodels && has_tidytext, fig.width=7, fig.height=4}
plot_top_terms(top_terms)
```

`plot_dtw()` shows the distribution of document-topic weights. This is useful
for quickly seeing whether topics are sharply assigned or diffuse across
documents.

```{r plot-dtw, fig.width=7, fig.height=4}
plot_dtw(fit)
```

`get_representative_candidates()` ranks documents by their dominant topic
weight. It can include stored metadata and text when those were supplied to
`fit_topic_model()` or passed explicitly.

```{r representatives}
representatives <- get_representative_candidates(
  fit,
  doc_data = docs,
  include_text = TRUE
)

representatives[
  ,
  .(doc_id, topic_max_id, topic_max_value, candidate_band, topic_rank, text)
]
```

## Predict New Documents

`predict_topic_model()` applies a fitted topic model to new documents. The
prediction input is aligned to the fitted vocabulary before backend prediction,
and the returned DTW table uses the same `Topic###` naming contract.

```{r predict}
new_docs <- data.frame(
  doc_id = paste0("new", 1:3),
  text = c(
    "Recurring revenue improved after subscription renewals.",
    "Audit remediation and compliance controls remained a priority.",
    "Debt refinancing lowered liquidity risk and interest expense."
  ),
  source = c("forecast", "governance", "capital"),
  stringsAsFactors = FALSE
)

new_corp <- quanteda::corpus(new_docs, text_field = "text", docid_field = "doc_id")
new_toks <- quanteda::tokens(new_corp, remove_punct = TRUE)
new_toks <- quanteda::tokens_tolower(new_toks)
new_toks <- quanteda::tokens_remove(new_toks, pattern = quanteda::stopwords("en"))
new_dfm <- quanteda::dfm(new_toks)

pred <- predict_topic_model(
  fit,
  new_dfm,
  doc_data = new_docs,
  include_text = TRUE
)

pred
```

## Evaluate Model Quality

`evaluate_topic_model()` reports quality metrics in one long-format table.
Metrics can be returned at aggregate level, topic level, or both.

```{r evaluate}
evaluation <- evaluate_topic_model(
  fit,
  training = dfm,
  metrics = c("coherence_umass", "diversity", "exclusivity", "train_perplexity"),
  top_n = 5L,
  level = "all"
)

evaluation
```

## Select K

`select_k_topics()` fits and evaluates a grid of candidate topic counts. Its
default behavior is unchanged when stability arguments are omitted. When
`stability_seeds` is supplied, aggregate stability is added to the selection
table and the full stability output is attached as an attribute.

```{r select-k}
k_selection <- select_k_topics(
  dfm,
  engine = "topicmodels",
  model = "lda",
  method = "Gibbs",
  k_grid = 2:3,
  metrics = c("coherence_umass", "diversity", "exclusivity"),
  top_n = 5L,
  holdout = 0,
  seed = 10L,
  stability_seeds = c(1L, 2L),
  control = list(fit = list(iter = 75L, burnin = 0L, thin = 1L))
)

k_selection
attr(k_selection, "stability")$k2
```

## Assess Stability Directly

Topic labels are arbitrary across repeated runs. A topic called `Topic001` in
one seed may correspond to `Topic002` in another seed. `assess_topic_stability()`
therefore repeatedly calls `fit_topic_model()`, extracts standardized TWW,
matches topics across runs, and reports matched-topic cosine similarities.

The automatic workflow changes only the seed unless `resampling` is explicitly
supplied.

```{r stability-auto}
stability <- assess_topic_stability(
  dfm,
  engine = "topicmodels",
  model = "lda",
  method = "Gibbs",
  k = 2,
  seeds = 1:3,
  control = list(fit = list(iter = 75L, burnin = 0L, thin = 1L))
)

stability
```

Users who prefer to fit models manually can pass a list of `nlp_topic_fit`
objects and still use package-owned topic matching and scoring.

```{r stability-manual}
fits <- lapply(1:3, function(seed) {
  fit_topic_model(
    dfm,
    engine = "topicmodels",
    model = "lda",
    method = "Gibbs",
    k = 2,
    control = list(fit = list(seed = seed, iter = 75L, burnin = 0L, thin = 1L))
  )
})

assess_topic_stability(fits, seeds = 1:3)
```

## Summarize and Export

`summarize_topics()` builds the interpretation table many research workflows
need: top terms, top-term probabilities, prevalence, available metrics,
representative documents, and optional text or metadata.

```{r summarize}
topic_summary <- summarize_topics(
  fit,
  training = dfm,
  doc_data = docs,
  top_n = 5L,
  representative_n = 2L,
  include_text = TRUE
)

topic_summary
```

The `representative_documents` column is a list column containing metadata-rich
document tables. Drop it for a flat export, or keep it when working inside R.

```{r export}
export_summary <- data.table::copy(topic_summary)
export_summary[, representative_documents := NULL]
data.table::fwrite(export_summary, tempfile(fileext = ".csv"))
```

## Backend Portability

The core interface is the same across supported backends. The examples below
are not evaluated in this vignette because optional backend availability varies
across machines and CI environments.

```{r text2vec-backend, eval=FALSE}
fit_text2vec <- fit_topic_model(
  dfm,
  engine = "text2vec",
  model = "lda",
  k = 2,
  control = list(fit = list(n_iter = 100, progressbar = FALSE))
)

get_top_terms(fit_text2vec, n = 5)
evaluate_topic_model(fit_text2vec, training = dfm, metrics = c("diversity", "exclusivity"))
```

```{r seededlda-backend, eval=FALSE}
fit_seededlda <- fit_topic_model(
  dfm,
  engine = "seededlda",
  model = "lda",
  k = 2
)

get_dtw(fit_seededlda)
get_tww(fit_seededlda)
```

Seeded LDA uses the same wrapper but requires a dictionary.

```{r seededlda-dictionary, eval=FALSE}
seed_dict <- quanteda::dictionary(list(
  performance = c("revenue", "subscription", "margin"),
  governance = c("audit", "control", "compliance")
))

fit_seeded <- fit_topic_model(
  dfm,
  engine = "seededlda",
  model = "seededlda",
  dictionary = seed_dict
)

summarize_topics(fit_seeded, doc_data = docs)
```

## Optional ETM Extensions

Embedded topic models are available when **topicmodels.etm**, **torch**, and a
working torch backend are installed. ETM fits use the same downstream accessors
for DTW, TWW, top terms, prediction, evaluation, and summaries. They also expose
ETM-specific topic and term embeddings.

```{r etm-backend, eval=FALSE}
fit_etm <- fit_topic_model(
  dfm,
  engine = "topicmodels.etm",
  model = "etm",
  k = 2,
  control = list(
    model = list(embeddings = 50L),
    fit = list(epochs = 10L, batch_size = 8L)
  )
)

get_topic_embeddings(fit_etm)
get_term_embeddings(fit_etm)
plot_topic_embeddings(fit_etm, top_n = 10)
```

## API Map

The main topic-model API is intentionally layered:

| Task | Primary function |
|---|---|
| Fit one model | `fit_topic_model()` |
| Extract DTW/TWW | `get_dtw()`, `get_tww()` |
| Interpret terms and documents | `get_top_terms()`, `get_representative_candidates()` |
| Plot standardized outputs | `plot_top_terms()`, `plot_dtw()` |
| Predict new documents | `predict_topic_model()` |
| Evaluate quality | `evaluate_topic_model()` |
| Select K | `select_k_topics()` |
| Assess seed stability | `assess_topic_stability()` |
| Build interpretation tables | `summarize_topics()` |
| Use ETM embeddings | `get_topic_embeddings()`, `get_term_embeddings()`, `plot_topic_embeddings()` |

